---
title: "Introduction to cardiometR"
format: html
vignette: >
  %\VignetteIndexEntry{Introduction to cardiometR}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**cardiometR** is an R package for analyzing Cardiopulmonary Exercise Testing (CPET) data. It provides:

- Import functions for COSMED Quark CPET metabolic cart exports
- Type-safe S7 classes for CPET data structures
- Peak value detection and ventilatory threshold analysis
- 9-panel CPET visualization following clinical guidelines
- Bilingual (English/French) PDF report generation
- Interactive Shiny application for clinical use

## Installation

```{r}
#| label: install
#| eval: false
# Install from GitHub
# install.packages("pak")
pak::pak("jotremblay/cardiometR")
```

## Quick Start

### Loading Data

```{r}
#| label: load-data
#| eval: false
library(cardiometR)

# Import COSMED xlsx file
data <- read_cpet("path/to/cosmed_export.xlsx")

# View participant information
print(data@participant)

# View test metadata
print(data@metadata)

# Preview breath-by-breath data
head(data@breaths)
```

`read_cpet()` works out what it is reading from the contents of the file, so an
export in French, or one configured with a different set of columns, needs no
extra arguments. Values are converted into the units the rest of the package
expects, and phase labels are translated to `rest`, `warmup`, `exercise` and
`recovery` whatever language the file uses.

Run interactively, it prints what it recognised, what it converted, and what it
left out. That summary is kept on the object:

```{r}
#| label: import-report
#| eval: false
report <- cpet_import_report(data)

# Which source column became which internal one, and in what units
report@columns[report@columns$status == "mapped", ]

# Anything it could not place
report@unknown
```

### When a Column Is Not Recognised

`preview_cpet_columns()` reads only the top of a file and reports how each
column would be interpreted. It works even on a file that `read_cpet()`
refuses, so it is the place to start when an import fails or a variable you
expected is missing.

```{r}
#| label: preview-columns
#| eval: false
preview_cpet_columns("path/to/export.xlsx")
```

The `source` column gives each header exactly as the file spells it. Pass those
spellings back if you need to name a column yourself:

```{r}
#| label: manual-mapping
#| eval: false
data <- read_cpet(
  "path/to/export.xlsx",
  mapping = c("Debit ventilatoire" = "ve_l"),
  units   = c(vo2_ml = "L/min")
)
```

### Adding Support for Your Own Metabolic Cart

Formats are described by YAML files rather than by R code, so a cart the
package has never seen can be added without changing the package. Start from
one of the packaged descriptions:

```{r}
#| label: list-dialects
#| eval: false
list_cpet_dialects()
```

Copy the closest one into your own dialect directory, edit it, and tell
cardiometR where to look:

```{r}
#| label: user-dialect
#| eval: false
dir.create(file.path(tools::R_user_dir("cardiometR", "config"), "dialects"),
           recursive = TRUE)

# ... write my-cart.yml there ...

clear_dialect_cache()
list_cpet_dialects()
```

A description needs surprisingly little. Everything general, including the
English and French names for each variable and the phase vocabulary, is
inherited from the shared `_core.yml`, so a new file only declares what is
peculiar to that cart:

``` yaml
version: 1
name: mycart
label: "Acme Cart 3000"
extends: _core
extensions: [csv]

detect:
  - where: header
    matches: "^AcmeTime$"
    weight: 9

aliases:
  time_s: ["acmetime"]
  vo2_ml: ["acme o2"]
```

You can also point at a directory with the `cardiometR.dialect_dir` option or
the `CARDIOMETR_DIALECTS` environment variable. A file whose name matches a
packaged one replaces it, which is the way to adjust a supported cart for a
local quirk without waiting for a release.

### Basic Analysis Workflow

```{r}
#| label: analysis-workflow
#| eval: false

# 1. Validate the data
validation <- validate(data)
print(validation)

# 2. Apply rolling average (30-second window)
averaged <- average(data, method = "rolling", window = 30)

# 3. Find peak values
peaks <- find_peaks(averaged, averaging = 30)
print(peaks)

# 4. Detect ventilatory thresholds
thresholds <- detect_thresholds(averaged, methods = c("V-slope", "VE_VO2"))
print(thresholds)

# 5. Create analysis object
analysis <- CpetAnalysis(
  data = averaged,
  peaks = peaks,
  thresholds = thresholds,
  validation = validation
)
```

## S7 Class Structure

cardiometR uses S7 classes for type-safe data handling:

| Class | Description |
|-------|-------------|
| `Participant` | Patient demographics (ID, name, age, sex, height, weight) |
| `CpetMetadata` | Test information (date, device, protocol, conditions) |
| `CpetData` | Container for breath-by-breath data with participant/metadata |
| `PeakValues` | Maximal values (VO2peak, HRmax, VEmax, etc.) |
| `Thresholds` | Ventilatory thresholds (VT1, VT2) with detection methods |
| `ValidationReport` | Data quality checks with errors/warnings/info |
| `CpetAnalysis` | Complete analysis combining all components |
| `ReportConfig` | PDF report configuration options |

### Creating Objects Manually

```{r}
#| label: manual-objects
#| eval: false

# Create a participant
participant <- Participant(
  id = "P001",
  name = "Jean Dupont",
  age = 35,
  sex = "M",
  height_cm = 175,
  weight_kg = 72,
  sport = "cycling"
)

# Access properties with @
participant@age
participant@weight_kg
```

## Visualization

### 9-Panel CPET Display

The standard clinical CPET display shows key relationships:

```{r}
#| label: plot-panel
#| eval: false

# Create 9-panel plot
plot_cpet_panel(analysis, language = "en")

# French labels
plot_cpet_panel(analysis, language = "fr")
```

The panels include:

1. O2 Pulse vs Power - stroke volume response
2. VO2 vs Power - linear relationship (efficiency)
3. VE vs VCO2 - ventilatory efficiency slope
4. V-Slope (VCO2 vs VO2) - threshold detection
5. Ventilatory Equivalents (VE/VO2, VE/VCO2)
6. RER vs VO2 - substrate utilization
7. PETO2/PETCO2 vs VO2 - end-tidal gases
8. HR vs Power - cardiovascular response
9. VE vs Time - ventilatory pattern

### Individual Plots

```{r}
#| label: individual-plots
#| eval: false

# V-slope for threshold visualization
plot_v_slope(analysis, thresholds = thresholds)

# Ventilatory equivalents
plot_ventilatory_equivalents(analysis, x_axis = "vo2")

# Heart rate response
plot_heart_rate(data, x_axis = "time", show_zones = TRUE)

# Power output with VO2 overlay
plot_power(data, show_vo2 = TRUE)
```

### Normative Comparison

Compare measured values against predicted and sport-specific norms:

```{r}
#| label: normative
#| eval: false

# General population comparison
plot_predicted_comparison(analysis)

# Elite cyclist comparison (includes gross efficiency)
plot_predicted_comparison(analysis, sport = "cycling", level = "elite")

# Recreational runner comparison
plot_predicted_comparison(analysis, sport = "running", level = "recreational")
```

## PDF Report Generation

Generate professional bilingual reports using Typst:

```{r}
#| label: report
#| eval: false

# Configure report
config <- ReportConfig(
  language = "fr",
  institution = "Centre ÉPIC",
  technician = "Dr. Smith"
)

# Generate PDF
generate_report(
  analysis = analysis,
  output_file = "cpet_report.pdf",
  config = config,
  include_graphs = TRUE,
  athlete_sport = "cycling",
  athlete_level = "competitive"
)
```

### Using Institution Logos

```{r}
#| label: logos
#| eval: false

# Built-in logos
config <- ReportConfig(
  institution = "Université de Montréal",
  logo_path = get_logo("udem")
)

# Or custom logo
config <- ReportConfig(
  institution = "My Lab",
  logo_path = "path/to/logo.png"
)
```

## Interactive Shiny Application

Launch the interactive analysis application:

```{r}
#| label: shiny
#| eval: false

# Start the app
run_app()

# Start with French interface
run_app(language = "fr")
```

The app provides:

- **Upload**: Drag-and-drop COSMED file import with validation
- **Configure**: Edit participant info and analysis settings
- **Results**: Interactive tables and plots with threshold markers
- **Report**: Configure and download PDF reports

## Bilingual Support

All user-facing text supports English and French:

```{r}
#| label: i18n
#| eval: false

# Get translated label
tr("peak_values", "en")
#> "Peak Values"

tr("peak_values", "fr")
#> "Valeurs Maximales"
```

## Analysis Methods

### Peak Value Detection

Peak values use rolling averages (default 30 seconds) per ATS/ACCP guidelines:

```{r}
#| label: peaks-detail
#| eval: false

peaks <- find_peaks(data, averaging = 30)

# Access individual peaks
peaks@vo2_peak      # Absolute VO2 (mL/min)
peaks@vo2_kg_peak   # Relative VO2 (mL/kg/min)
peaks@hr_peak       # Heart rate (bpm)
peaks@ve_peak       # Minute ventilation (L/min)
peaks@rer_peak      # Respiratory exchange ratio
peaks@power_peak    # Power output (W)
```

### Ventilatory Threshold Detection

Multiple methods available for threshold detection:

```{r}
#| label: thresholds-detail
#| eval: false

thresholds <- detect_thresholds(
  data,
  methods = c("V-slope", "VE_VO2", "VE_VCO2", "PETO2")
)

# VT1 (aerobic threshold)
thresholds@vt1_vo2     # VO2 at VT1
thresholds@vt1_hr      # HR at VT1
thresholds@vt1_power   # Power at VT1
thresholds@vt1_method  # Detection method used

# VT2 (respiratory compensation point)
thresholds@vt2_vo2
thresholds@vt2_hr
thresholds@vt2_power
thresholds@vt2_method

# Detection confidence
thresholds@confidence  # "high", "moderate", "low", or "unable"
```

### Data Averaging

Three averaging methods available:

```{r}
#| label: averaging
#| eval: false

# Time-based bins (e.g., 30-second epochs)
avg_time <- average(data, method = "time", window = 30)

# Breath-based bins (e.g., every 10 breaths)
avg_breath <- average(data, method = "breath", window = 10)

# Rolling/moving average
avg_rolling <- average(data, method = "rolling", window = 30)
```

## Data Validation

Comprehensive validation following ATS/ACCP guidelines:

```{r}
#| label: validation
#| eval: false

validation <- validate(data)

# Check if data passed validation
validation@is_valid

# View any errors
validation@errors

# View warnings
validation@warnings

# View informational messages
validation@info
```

Validation checks include:

- Required columns present
- Physiological plausibility (VO2, HR, RER ranges)
- Time series monotonicity
- Data completeness
- Aberrant breath detection

## References

- ATS/ACCP Statement on Cardiopulmonary Exercise Testing. *Am J Respir Crit Care Med*. 2003;167(2):211-277.
- Beaver WL, Wasserman K, Whipp BJ. A new method for detecting anaerobic threshold by gas exchange. *J Appl Physiol*. 1986;60(6):2020-2027.
- Jones NL, et al. *Clinical Exercise Testing*. 4th ed. Saunders; 1997.

## See Also

- [spiro package](https://docs.ropensci.org/spiro/) - rOpenSci CPET package
- [whippr package](https://fmmattioni.github.io/whippr/) - VO2 kinetics analysis
