---
title: "Generating PDF Reports"
format: html
vignette: >
  %\VignetteIndexEntry{Generating PDF Reports}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`cardiometR` uses [Typst](https://typst.app) via the
[`typr`](https://cran.r-project.org/package=typr) package to compile
clinical PDF reports. Reports are fully bilingual (English / French) and
can include normative comparisons against FRIEND and CHEER registries.

## Minimal report

```{r}
library(cardiometR)

file     <- system.file("extdata", "example_cosmed.xlsx", package = "cardiometR")
data     <- read_cpet(file)
analysis <- CpetAnalysis(
  data       = average(data, method = "rolling", window = 30),
  peaks      = find_peaks(average(data)),
  thresholds = detect_thresholds(average(data))
)

config <- ReportConfig(language = "en")

generate_report(analysis, output_file = "cpet_report.pdf", config = config)
```

## ReportConfig options

```{r}
config <- ReportConfig(
  language          = "fr",                            # "en" or "fr"
  institution       = "Universite de Montreal",
  lab_name          = "LPEBA",
  lab_url           = "https://lpeba.umontreal.ca",
  logo_path         = get_logo("udem"),                # or any file path
  technician        = "Dr. Smith",
  prediction_source = "jones"                          # "jones" or "prefaut"
)
```

Built-in logos are available for `"udem"` and `"epic"` via `get_logo()`.
Pass any image path for a custom logo.

## Normative comparison in reports

Adding `athlete_sport` and `athlete_level` enables a side-by-side
comparison against the relevant endurance-athlete registry:

```{r}
# Elite cyclist
generate_report(analysis, "cyclist_report.pdf", config,
                athlete_sport = "cycling", athlete_level = "elite")

# Recreational runner
generate_report(analysis, "runner_report.pdf", config,
                athlete_sport = "running", athlete_level = "recreational")
```

Valid combinations: `sport ∈ {"cycling", "running", "triathlon"}`,
`level ∈ {"elite", "competitive", "recreational"}`.

## Selecting sections

When you only need a subset of the report, pass `report_sections`:

```{r}
generate_report(analysis, "short.pdf", config,
                report_sections = c("protocol_details", "thresholds", "graphs"))
```

Any section with missing data is omitted silently.

## Clinical notes and interpretation

`clinical_notes` accepts plain text; `interpretation` accepts a list of
named interpretation sections that the template renders verbatim.

```{r}
generate_report(analysis, "full.pdf", config,
                clinical_notes = "No symptoms during exercise.",
                interpretation = list(
                  aerobic   = "Maximal effort achieved (RER 1.18).",
                  cardiovascular = "Normal chronotropic response."
                ))
```

## Bilingual labels

Report labels live in `inst/translations/labels_{en,fr}.yml`. To add a
new label, add it to both files (the test suite enforces key parity),
then reference it from the Typst template via the `tr()` helper.
