---
title: "S7 Classes"
format: html
vignette: >
  %\VignetteIndexEntry{S7 Classes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

cardiometR uses [S7](https://rconsortium.github.io/S7/) classes so that every
piece of CPET data carries a validated type. Properties are accessed with
the `@` operator — never `$`.

## Class hierarchy

| Class | Role | Key properties |
|-------|------|----------------|
| `Participant` | Patient demographics | `id`, `age`, `sex`, `height_cm`, `weight_kg`, `sport` |
| `CpetMetadata` | Test information | `test_date`, `device`, `protocol`, `modality` |
| `CpetData` | Breath-by-breath container | `breaths`, `participant`, `metadata` |
| `PeakValues` | Maximal values | `vo2_peak`, `vo2_kg_peak`, `hr_peak`, `rer_peak`, `ve_peak` |
| `Thresholds` | Ventilatory thresholds | `vt1`, `vt2`, `methods_used` |
| `CpetAnalysis` | Combined analysis object | `data`, `peaks`, `thresholds`, `quality` |
| `ReportConfig` | PDF report settings | `language`, `institution`, `lab_name`, `logo_path` |

Quality assessment classes (`ExerciseQualityCriteria`, `ProtocolQuality`,
`DataQualityReport`, `QualityAssessment`) live alongside and are populated
by `assess_quality()`.

## Loading data into S7 objects

```{r}
#| eval: false
library(cardiometR)

file <- system.file("extdata", "example_cosmed.xlsx", package = "cardiometR")
data <- read_cpet(file)

class(data)
#> <cardiometR::CpetData>

data@participant@age
data@metadata@test_date
nrow(data@breaths)
```

## Chaining into an analysis

```{r}
#| eval: false
averaged <- average(data, method = "rolling", window = 30)
peaks    <- find_peaks(averaged)
thr      <- detect_thresholds(averaged)
quality  <- assess_quality(data, rpe = 18, lactate = 9.5)

analysis <- CpetAnalysis(
  data       = averaged,
  peaks      = peaks,
  thresholds = thr,
  quality    = quality
)

analysis@peaks@vo2_kg_peak
analysis@thresholds@vt1
analysis@quality@overall_grade
```

## Why S7?

* **Validation on construction** — invalid data (e.g., age outside a
  plausible range) fails at the constructor, not deep inside analysis
  code.
* **Generic dispatch** — `plot()`, `print()`, `validate()`, `average()`,
  `find_peaks()`, `detect_thresholds()` all dispatch on class, so the
  same verbs work on raw `CpetData` and on `CpetAnalysis`.
* **Forward compatibility** — S7 is the R Consortium successor to S4 and
  R6 for heterogeneous, typed data.

See `?S7::new_class` and `vignette("S7", package = "S7")` for the class
system itself.
